---
title: "Continuous-Time-SEM"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{CTSEM}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, include=FALSE}
library(lessTemplates)
library(lavaan)
library(lessSEM)
library(ctsemOMX)
set.seed(123)
```

**lessTemplates** also allows for implementing continuous time structural 
euqation models (CTSEM; Voelkle et al., 2012; Driver et al., 2017). This model
allows for unequally spaced measurement occasions in longitudinal data sets by
relying on the assumption that the dyanmical processes can be explained using stochastic
differential equations. The model is given by

\begin{equation}
\text d\pmb \eta(t) = \pmb A \pmb \eta(t)\text dt + \pmb G \text d \pmb W(t).
\label{eqn:ctsem}
\end{equation}

$\pmb \eta(t)$ is a vector of latent states at time point $t$, $\pmb A$ is the
drift-matrix with auto-effects in the diagonal and cross-effects in the off-diagonal, 
and $\pmb W(t)$ is a Wiener process vector. $\pmb G$ is the diffusion matrix which
influences the variance-covariance matrix of the latent residuals. A thorough
introduction is, for instance, provided by Voelkle et al. (2012).

Please note that the R packages [**ctsem**](https://github.com/driveraus/ctsem) 
and [**ctsemOMX**](https://github.com/driveraus/ctsemOMX) (Driver et al., 2017)
provide more stable implementations of CTSEM and [**regCtsem**](https://github.com/jhorzek/regCtsem)
is a specialized package which allows for regularization of such models (Orzek & Voelkle, In Press).
In practice, you may find that **ctsem** and **regCtsem** are faster than the implementation
in **lessTemplates**. The objective of the CTSEM implementation in **lessTemplates**
is to show the versatility of **lessSEM**, not to provide an alternative to
existing implementations.

## The Data Set

**lessTemplates** expects the data to be given in long format, with a column
named `person` and a column named `time`. We will use the `AnomAuth` data set in
the following (Voelkle et al., 2012; Driver et al., 2017):
```{r}
library(ctsemOMX)
data("AnomAuth")
head(AnomAuth)
```
Note that the data is in long format, not in wide format. We can transform the
data set using **ctsemOMX**:

```{r}
# from long to wide
data <- ctWideToLong(datawide = AnomAuth,
                     Tpoints= 5,
                     n.manifest=2,
                     manifestNames = c("Y1", "Y2"))
data <- ctDeintervalise(datalong = data, 
                        id='id', 
                        dT='dT')
colnames(data) <- c("person", "time", "Y1", "Y2")
data <- as.data.frame(data)
data <- data[!(is.na(data$Y1) & is.na(data$Y2)),]
head(data)
```

## Specifying the CTSEM

The model syntax is similar to the mathematical notation shown in the equations above.

```{r}
model <- "
# Specify the latent dynamics in differential equation model notation:
d_eta1(t) ~ eta1(t) + eta2(t)
d_eta2(t) ~ eta1(t) + eta2(t)

# Covariances (Wiener process)
d_eta1(t) ~~ d_eta1(t) + d_eta2(t)
d_eta2(t) ~~ d_eta2(t)

# Latent intercepts
eta1(0) ~ 1
eta2(0) ~ 1

# Measurement model
eta1(t) =~ 1*Y1(t)
eta2(t) =~ 1*Y2(t)

# Manifest variances and covariances
Y1(t) ~~ 0*Y1(t) + 0*Y2(t)
Y2(t) ~~ 0*Y2(t)

# Manifest intercepts
Y1(t) ~ m1*1
Y2(t) ~ m2*1
"
```
Note the special notation:

1. A variable starting with `d_` is assumed to be a differential that is predicted
using the current states of the latent variables.
2. If a variable is time-dependent, this is denoted with the special operator `(t)`.

Currently, we can only use latent, time-dependent variables as predictors in the 
dynamical equations. The following is not allowed: `d_eta1(t) ~ eta1(t) + kappa`.
This is because `kappa` is not time-dependent.

## Using `CTSEM` to build the model

CTSEM is implemented in **lessTemplates** using the transformations of **lessSEM**
(see `vignette(topic = "Parameter-transformations", package = "lessSEM")` for more
details).

```{r}
ctsem <- CTSEM(model = model, 
               data = data)
names(ctsem)
```
The created object has two elements of relevance in the following: A lavaan object
(`lavaanModel`) and the transformations necessary to build a CTSEM from this lavaan 
object (`transformation`). 

<details>
<summary>Show lavaan model details</summary>
```{r}
summary(ctsem$lavaanModel)
```
</details>

<details>
<summary>Show transformation details</summary>
```{r}
cat(ctsem$transformation)
```
</details>

## Fitting the model with **lessSEM**

To estimate the model, we can use any of the optimizers within **lessSEM**. If
no regularization is required, we can use the `bfgs` function:
```{r,include=FALSE}
library(lessSEM)

fit <- bfgs(lavaanModel = ctsem$lavaanModel, 
            modifyModel = modifyModel(transformations = ctsem$transformation))
```
```{r,eval=FALSE}
library(lessSEM)

fit <- bfgs(lavaanModel = ctsem$lavaanModel, 
            modifyModel = modifyModel(transformations = ctsem$transformation))
```
The parameter estimates are given by
```{r}
coef(fit)
```
Note that `CTSEM` automatically estimates the initial covariances and means separately
from all other covariances and means.

<details>
<summary>Compare to **ctsemOMX**</summary>
The following is copied directly from `?ctsemOMX::ctFit`
```{r}
AnomAuthmodel <- ctModel(LAMBDA = matrix(c(1, 0, 0, 1), 
                                         nrow = 2, 
                                         ncol = 2), 
                         Tpoints = 5, 
                         n.latent = 2, 
                         n.manifest = 2, 
                         MANIFESTVAR=diag(0, 2), 
                         TRAITVAR = NULL) 
AnomAuthfit <- ctFit(AnomAuth, 
                     AnomAuthmodel)
summary(AnomAuthfit)
```
</details>

## Adding a Random Intercept

Similar to the RI-CLPM (Hamaker et al., 2015), we can specify a random intercept in CTSEM as well
(Driver et al., 2017):

```{r}
modelRI <- "
# Specify the latent dynamics in differential equation model notation:
d_eta1(t) ~ eta1(t) + eta2(t)
d_eta2(t) ~ eta1(t) + eta2(t)

# Covariances (Wiener process)
d_eta1(t) ~~ d_eta1(t) + d_eta2(t)
d_eta2(t) ~~ d_eta2(t)

# Latent intercepts
eta1(0) ~ 1
eta2(0) ~ 1

# Measurement model
eta1(t) =~ 1*Y1(t)
eta2(t) =~ 1*Y2(t)

# Manifest variances and covariances
Y1(t) ~~ 0*Y1(t) + 0*Y2(t)
Y2(t) ~~ 0*Y2(t)

# Manifest intercepts
Y1(t) ~ m1*1
Y2(t) ~ m2*1

# Add random intercepts
RI1 =~ 1*Y1(t)
RI2 =~ 1*Y2(t)

RI1 ~~ RI1 + RI2
RI2 ~~ RI2
"
```

```{r,include=FALSE}
ctsemRI <- CTSEM(model = modelRI, 
                 data = data)
fitRI <- bfgs(lavaanModel = ctsemRI$lavaanModel, 
              modifyModel = modifyModel(transformations = ctsemRI$transformation))
```
```{r,eval=FALSE}
ctsemRI <- CTSEM(model = modelRI, 
                 data = data)
fitRI <- bfgs(lavaanModel = ctsemRI$lavaanModel, 
              modifyModel = modifyModel(transformations = ctsemRI$transformation))
```
The parameter estimates are given by
```{r}
coef(fitRI)
```
We can compare the model with and the model without random intercepts using
the information criteria implemented in **lessSEM**:

```{r}
AIC(fit)
AIC(fitRI)
```
Note that the random-intercept CTSEM outperforms the CTSEM in terms of the AIC.

<details>
<summary>Compare to **ctsemOMX**</summary>
The following is copied and adapted directly from `?ctsemOMX::ctFit`
```{r}
AnomAuthmodel <- ctModel(LAMBDA = matrix(c(1, 0, 0, 1), 
                                         nrow = 2, 
                                         ncol = 2), 
                         Tpoints = 5, 
                         n.latent = 2, 
                         n.manifest = 2, 
                         MANIFESTVAR=diag(0, 2), 
                         TRAITVAR = NULL,
                         MANIFESTTRAITVAR = "auto" # add random intercepts
                         ) 
AnomAuthfit <- ctFit(AnomAuth, 
                     AnomAuthmodel)
summary(AnomAuthfit)
```
</details>

## Bibliography

- Driver, C. C., Oud, J. H. L., & Voelkle, M. C. (2017). Continuous Time Structural Equation Modelling With R Package ctsem. Journal of Statistical Software, 77(5), 1–36. https://doi.org/10.18637/jss.v077.i05
- Hamaker, E. L., Kuiper, R. M., & Grasman, R. P. P. P. (2015). A critique of the cross-lagged panel model. Psychological Methods, 20(1), 102–116. https://doi.org/10.1037/a0038889
- Orzek, J. H., & Voelkle, M. C. (In Press). Regularized Continuous Time Dynamic Networks. Psychological Methods.
- Voelkle, M. C., Oud, J. H. L., Davidov, E., & Schmidt, P. (2012). An SEM Approach to Continuous Time Modeling of Panel Data: Relating Authoritarianism and Anomia. Psychological Methods, 17(2), 176–192. https://doi.org/10.1037/a0027543
